---
title: "Diffusive equilibration in thin-films (DET) - helper functions"
author: "Marko Lipka"
date: "February 26, 2018"
output: github_document
bibliography: library.bib
---

Collection of simple functions for calculations in the context of DET applications:

  * *DET.slice.volume.ccm(slice.width.mm, gel.thickness.mm = 1, DET.window.width.mm = 20)* calculates the volume of a rectangular DET gel slice from 
    + DET window width (default = 20\ mm),
    + gel thickness (default = 1\ mm) and
    + slice width

  * *DET.slice.amount.uM(V.slice.ccm, c.ambient.uM, ...)* returns the amount of substance of a solute in a slice of given volume (*V.slice.ccm*) or geometry (arguments passed to *DET.slice.volume.ccm()*, see above) after equilibration with ambient solute concentration

  * *DET.extract.conc.uM(n.slice.umol, V.extraction.uL, ...)* returns the concentration of a gel extract made with a given eluent volume and the amount of substance of a solute in the extracted slice.
  Instead of *n.slice.umol*, arguments of the above described functions can be provided instead (handed to the *DET.slice.amount.uM()* function).

# Background

Biogeochemical processes in organic-rich sediments or fluxes through the benthic boundary layer are reflected by concentration gradients of nutrients, redox-metabolites and metals.
It is desirable to sample these transition zones at highest possible resolutions.
The diffusive equilibration in thin-films (DET) technique is based on the application of thin polyacrylamide gel films, which are convenient to prepare, stable, easy and clean to use, and require a relatively short time for solution equilibration (within hours) [@Krom1994; @Harper1997].
The DET gels are usually sliced to the required spatial resolution after recovery.
For concentration analysis, the gel slices are back equilibrated by extracting them with a small volume of deionized water or another eluent.
To minimize back-equilibration times for DET analysis, diffusional path lengths and eluent volumes should be as small as possible [@Harper1997].

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

```{r test}
library(testthat)
test_file("tests/test_DET-helper.R",
          reporter = "fail") # Stops on failed unit-test, otherwise silent
```

# Examples

```{r exampleSetup, include = F}
library(reshape2)
library(RSQLite)

DBfile <- "~/Dropbox/IOW/SECOS/data/DB/SECOS-DB.sqlite"
drv <- dbDriver("SQLite")
con <- dbConnect(drv, DBfile)

source("~/Dropbox/IOW/R-functions/DET_helpers/DET_helpers.R")
```

## DET application in marine sediments
### Known pore water concentration profiles from sampling with Rhizons

```{r loadandplot, echo=FALSE, fig.width=10}
pw.POS475.DET <- dbGetQuery(con, "SELECT * FROM Porewater_analysis_results_SECOS
                       NATURAL LEFT JOIN cruises
                       WHERE Cruises_ID = 'POS475'
                       AND Stations_ID IN (53, 57);")

mains <-  c("Ca_mM",  "Mg_mM",  "Na_mM",  "K_mM",  "SO4_mM")
minors <- c("Ba_µM",  "Li_µM",  "Fe_µM",  "Mn_µM",  "P_µM",  "Si_µM",  "Sr_µM",  "Mo_µM")

library(tidyverse)
ggplot(melt(pw.POS475.DET, measure.vars = c(mains, minors)), aes(x = Depth_cm, y = value)) +
  geom_point() +
  coord_flip() +
  scale_x_reverse() +
  expand_limits(y = 0) +
  facet_grid(Site ~ variable, scales = "free", switch = "x") +
  theme_bw() +
  theme(strip.placement = "outside", axis.text.x = element_text(angle = 45, hjust = 1))
```

### Potential range of concentration values in DET slices applied in the same sediments

```{r}
slice.width.cm <- 0.1
extraction.volume.uL <- 100

slice.volume.ccm <- DET.slice.volume.ccm(slice.width.mm = slice.width.cm * 10)
```

With a DET gel slice width of `r slice.width.cm`\ cm
and an extraction volume of `r extraction.volume.uL`\ µL, the following concentration ranges can be expected in the DET slice extracts.

```{r, echo = F}
pw.POS475.DET.slice.n.uM <- 
  data.frame(pw.POS475.DET[1:5],
             lapply(pw.POS475.DET[, mains],
                    function(x) {
                      DET.extract.conc.uM(
                        V.slice.ccm = slice.volume.ccm, 
                        c.ambient.uM = x * 1000,
                        V.extraction.uL = extraction.volume.uL
                      )} / 1000),
             lapply(pw.POS475.DET[, minors],
                    function(x) {
                      DET.extract.conc.uM(
                        V.slice.ccm = slice.volume.ccm,
                        c.ambient.uM = x,
                        V.extraction.uL = extraction.volume.uL)}))

pw.POS475.DET.slice.n.uM %>%
  select(mains, minors) %>%
  apply(MARGIN = 2, FUN = range) %>%
  t() %>%
  pander::pander(digits = 0)
```

Based on these values, the further analysis of the extracts can be planned precisely, so that only a minimum of extract volume must be used. 

# ToDo

  * calculate application times depending on gel thickness
  * get slope from highres conc profile
  * plot 2D solute distribution when gel cut into a matrix instead of slices

# References